---
title: "Mixture of Gaussians models and Expectation Maximization"
output: html_notebook
---

This is an exploration on using Expectation Maximization to estimate the parameters of a simple mixture of gaussians model in R . I followed Bishop (2006), pag 430 as a guide.  

```{r}
library(ggplot2)
```

## Gaussian Distribution



```{r}
set.seed(42)
qplot(rnorm(1000))

```

## Mixtures of Gaussians

A Gaussian mixture model is a simple linear superposition of Gaussian components, where a discrete latent variable controls what Gaussian distribution produces a given datapoint.


$$p(\mathbf{x}) = \sum_{k=1}^K \pi_{k} \mathcal{N}(\mathbf{x} \mid \mathbf{\mu_{k}},\mathbf{\Sigma_{k}} )  $$

The latent variable $\mathbf{z}$ is an indicator binary random variable, in which only a particular element $z_{k}$ is equal to 1 and all other elements are equal to 0. 


```{r}
n = 1000
p = c(1,1,1)
z = rmultinom(n,1,p)

```






```{r}

df = data.frame(
  a = rnorm(n,0,5), 
  b = rnorm(n,10,1),
  c = rnorm(n,20,5)
)

x = colSums(z * t(as.matrix(df)))

```

```{r}

data = data.frame(
  'x' = x, 
  'z' =sapply(1:n, function(x) which(z[,x] == 1) )   
)


ggplot(data, aes(x = x)  ) +
  geom_histogram( aes( fill = factor(z) ), binwidth = 1 ) +
  geom_density( aes(y=1 * ..count.., color = factor(z) )  ) 

```


```{r}

pi = c(0.1,0.4,0.5)
mu = c(0,5,25)
sigma = c(1,2,1)

```



```{r}

ggplot(data = data.frame(x = c(-20, 45)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean=mu[1], sd=sigma[1])) +
  stat_function(fun = dnorm, n = 101, args = list(mean=mu[2], sd=sigma[2])) +
  stat_function(fun = dnorm, n = 101, args = list(mean=mu[3], sd=sigma[3])) +
  scale_y_continuous()

```



```{r}

gamma = data.frame(
 '1' = pi[1] * dnorm(x,mu[1],sigma[1]), 
 '2' = pi[2] * dnorm(x,mu[2],sigma[2]),
 '3' = pi[3] * dnorm(x,mu[3],sigma[3])
)

total = rowSums(gamma)
gamma = gamma / total
assignments = max.col(gamma)

mu_t = c(sum(gamma[,1] * x) / sum(gamma[,1]),
         sum(gamma[,2] * x) / sum(gamma[,2]),
         sum(gamma[,3] * x) / sum(gamma[,3]))


sigma_t = c(
  sqrt(sum(gamma[,1] * (x - mu_t[1])^2) / sum(gamma[,1])),
  sqrt(sum(gamma[,2] * (x - mu_t[2])^2) / sum(gamma[,2])),
  sqrt(sum(gamma[,3] * (x - mu_t[3])^2) / sum(gamma[,3]))
)


pi_t = prop.table(table(assignments))

mu_t
sigma_t
pi_t

mu = mu_t
sigma = sigma_t
pi = as.vector(pi_t)

ggplot(data = data.frame(x = c(-20, 45)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean=mu[1], sd=sigma[1])) +
  stat_function(fun = dnorm, n = 101, args = list(mean=mu[2], sd=sigma[2])) +
  stat_function(fun = dnorm, n = 101, args = list(mean=mu[3], sd=sigma[3])) +
  scale_y_continuous()


```

```{r}

pi = c(0,0,1)

```