---
title: "Mixture of Gaussians models and Expectation Maximization"
output: html_notebook
---

This is an exploration on using Expectation Maximization to estimate the parameters of a simple mixture of gaussians model in R . I followed Bishop (2006), pag 430 as a guide.  

```{r}
library(ggplot2)
```

## Gaussian Distribution



```{r}
set.seed(42)
qplot(rnorm(1000))

```

## Mixtures of Gaussians

A Gaussian mixture model is a simple linear superposition of Gaussian components, where a discrete latent variable controls what Gaussian distribution produces a given datapoint.


$$p(\mathbf{x}) = \sum_{k=1}^K \pi_{k} \mathcal{N}(\mathbf{x} \mid \mathbf{\mu_{k}},\mathbf{\Sigma_{k}} )  $$

The latent variable $\mathbf{z}$ is an indicator binary random variable, in which only a particular element $z_{k}$ is equal to 1 and all other elements are equal to 0. 


```{r}
n = 1000
p = c(1,1,1)
z = rmultinom(n,1,p)

```






```{r}

df = data.frame(
  a = rnorm(n,0,5), 
  b = rnorm(n,10,1),
  c = rnorm(n,20,5)
)

x = colSums(z * t(as.matrix(df)))

```

```{r}

data = data.frame(
  'x' = x, 
  'z' =sapply(1:n, function(x) which(z[,x] == 1) )   
)


ggplot(data, aes(x = x)  ) +
  geom_histogram( aes( fill = factor(z) ), binwidth = 1 ) +
  geom_density( aes(y=1 * ..count.., color = factor(z) )  ) 

```


```{r}

pi = c(0.1,0.4,0.5)
mu = c(0,5,25)
sigma = c(1,2,1)

params = list('mu' = mu, 'sigma' = sigma, 'pi' = pi )  

```



```{r}


#ggplot(data = data.frame(x = c(-20, 45)), aes(x)) +

ggplot(data, aes(x = x)  ) +
  geom_histogram( aes( fill = factor(z) ), binwidth = 1 ) +
  geom_density( aes(y=1 * ..count.., color = factor(z) ) ) +
  stat_function(fun = function(x) n * pi[1] * dnorm(x, mean=mu[1], sd=sigma[1]), n = 101 ) +
  stat_function(fun = function(x) n * pi[2] * dnorm(x, mean=mu[2], sd=sigma[2]), n = 101 ) +
  stat_function(fun = function(x) n * pi[3] * dnorm(x, mean=mu[3], sd=sigma[3]), n = 101 ) +
  scale_y_continuous()

```


```{r}

loglikelihood <- function(x, pi,mu,sigma) {
  sum(log(rowSums(point_likelihood(x, pi,mu,sigma))))
}

point_likelihood <- function(x, pi,mu,sigma) {
 gamma = data.frame(
 '1' = pi[1] * dnorm(x,mu[1],sigma[1]), 
 '2' = pi[2] * dnorm(x,mu[2],sigma[2]),
 '3' = pi[3] * dnorm(x,mu[3],sigma[3])
  )

  gamma 
}


e_step <- function(x, pi,mu,sigma) {
  gamma = point_likelihood(x, pi,mu,sigma)
  total = rowSums(gamma)
  gamma = gamma / total
}

m_step <- function(x, gamma,pi,mu,sigma) {
  
  mu_t = c(sum(gamma[,1] * x) / sum(gamma[,1]),
           sum(gamma[,2] * x) / sum(gamma[,2]),
           sum(gamma[,3] * x) / sum(gamma[,3]))
  
  
  sigma_t = c(
    sqrt(sum(gamma[,1] * (x - mu_t[1])^2) / sum(gamma[,1])),
    sqrt(sum(gamma[,2] * (x - mu_t[2])^2) / sum(gamma[,2])),
    sqrt(sum(gamma[,3] * (x - mu_t[3])^2) / sum(gamma[,3]))
  )

  assignments = max.col(gamma)
  pi_t = prop.table(table(assignments))
  list('mu' = mu_t, 'sigma' = sigma_t, 'pi' = as.vector(pi_t) )  
}




```

```{r}


normparams_to_text <-function(pi,mu,sigma) {
  sprintf("mu: %2.2f \n sigma: %2.2f \n pi: %1.3f", mu, sigma, pi) 
}


plot_distribution <- function(data, step,  pi, mu, sigma) {

  ggplot(data, aes(x = x)  ) +
  geom_histogram( aes( fill = factor(z) ), binwidth = 1 ) +
  geom_density( aes(y=1 * ..count.., color = factor(z) ) ) +
  stat_function(fun = function(x) n * pi[1] * dnorm(x, mean=mu[1], sd=sigma[1]), n = 101 ) +
  stat_function(fun = function(x) n * pi[2] * dnorm(x, mean=mu[2], sd=sigma[2]), n = 101 ) +
  stat_function(fun = function(x) n * pi[3] * dnorm(x, mean=mu[3], sd=sigma[3]), n = 101 ) +
  geom_text(label = normparams_to_text(pi[1],mu[1],sigma[1]), x = mu[1], y = 150 ) +
  geom_text(label = normparams_to_text(pi[2],mu[2],sigma[2]), x = mu[2], y = 150 ) +
  geom_text(label = normparams_to_text(pi[3],mu[3],sigma[3]), x = mu[3], y = 150 ) +
  geom_text(label = sprintf("step %i",step) , x = -10, y = 150 ) +
  scale_y_continuous(limits = c(0,160))
  
}


```



```{r}

i <- 0
loglik_before <- loglikelihood(x,params$pi,params$mu,params$sigma)
  print(loglik_before)

plotlist = list()  


  
repeat {
  gamma  <- e_step(x, params$pi,params$mu,params$sigma)
  params <- m_step(x, gamma,params$pi,params$mu,params$sigma)
  loglik_after <- loglikelihood(x,params$pi,params$mu,params$sigma)
  print(loglik_after)
  
  if (i %% 5 == 0) {
    l <- length(plotlist) + 1
    plotlist[[l]] <- plot_distribution(data, i, params$pi,params$mu,params$sigma) 
  }
  
  if ( (loglik_after - loglik_before) < 0.01 ) {
    break
  } else {
    i <- i+1 
    loglik_before <- loglik_after
  }
  
}


```

```{r}

for (i in (1:length(plotlist))) {
  print(plotlist[[i]])
}

```

